Weighted Model Counting Without Parameter Variables

### 2021-02-15 doc/paper3/paper.tex
Fleshing out some lemmas

\paragraph{Notation.} For any propositional formula $\phi$ and $p, q \in
\mathbb{R}_{\ge 0}$, let $[\phi]^p_q\colon 2^X \to \mathbb{R}$ be a
pseudo-Boolean function defined as
\[
  [\phi]^p_q \coloneqq
  \begin{cases}
    p & \text{if } \phi \\
    q & \text{otherwise.}
  \end{cases}
\]

\paragraph{References.}
\begin{itemize}
\item Related work without publicly available implementations:
  \begin{itemize}
  \item direct compilation to SDDs \cite{DBLP:conf/ecsqaru/ChoiKD13}
  \item direct compilation to PSDDs, also eliminating parameter variables (a thesis)
  \item maybe two more papers
  \end{itemize}
\end{itemize}
% TODO: do a literature search focused around these papers
% TODO : use mathcal on V, L, E (and introduce them)
% TODO: maybe use texttt for DPMC?

@@

\paragraph{TODO: Define:}
\begin{itemize}
\item Formal definition of a previous WMC instance (CNF, literal weight
  function) and the new definition (set of $2^{X} \to \mathbb{R}_{\ge 0}$
  pseudo-Boolean functions and a constant). Note that the constant idea is
  borrowed from \texttt{bklm16}.
\item Boolean formula in CNF (perhaps this is too trivial to define),
\item primal graph of a CNF formula (a.k.a.
  Gaifman/(variable) interaction/connectivity/clique/representing graph),
\item Bayesian network (I already have a definition of these last two),
\item moralisation of a Bayesian network (or of any DAG),
\end{itemize}

@@

\begin{definition}[\cite{DBLP:conf/cp/DudekPV20}, verbatim]
  Let $X$ be a set of Boolean variables and $\phi$ be a CNF formula over $X$. A
  \emph{project-join tree} (PJT) of $\phi$ is a tuple $(T, r, \gamma, \pi)$
  where:
  \begin{itemize}
  \item $T$ is a tree with root $r \in \mathcal{V}(T)$,
  \item $\gamma\colon \mathcal{L}(T) \to \phi$ is a bijection between the leaves
    of $T$ and the clauses of $\phi$, and
  \item $\pi\colon \mathcal{V}(T) \setminus \mathcal{L}(T) \to 2^X$ is a
    labelling function on internal nodes.
  \end{itemize}
  Moreover, $(T, r, \gamma, \pi)$ must satisfy the following two properties:
  \begin{enumerate}
  \item $\{\pi(n) : n \in \mathcal{V}(T) \setminus \mathcal{L}(T)\}$ is a
    partition of $X$, and
  \item for each internal node $n \in \mathcal{V}(T) \setminus \mathcal{L}(T)$,
    variable $x \in \pi(n)$, and clause $c \in \phi$ such that $x$ appears in
    $c$, the leaf node $\gamma^{-1}(c)$ must be a descendant of $n$ in $T$.
  \end{enumerate}
\end{definition}

\begin{definition}[\cite{DBLP:conf/cp/DudekPV20}]
  \[
    \mathtt{Vars}(n) \coloneqq
    \begin{cases}
      \mathtt{Vars}(\gamma(n)) & \text{if } n \in \mathcal{L}(T) \\
      \left( \bigcup_{o \in \mathcal{C}(n)} \mathtt{Vars}(o) \right) \setminus
      \pi(n) & \text{if } n \not\in \mathcal{L}(T).
    \end{cases}
  \]
  \[
    \mathtt{size}(n) \coloneqq
    \begin{cases}
      |\mathtt{Vars}(n)| & \text{if } n \in \mathcal{L}(T) \\
      |\mathtt{Vars}(n) \cup \pi(n)| & \text{if } n \not\in \mathcal{L}(T).
    \end{cases}
  \]
  The \emph{width} of a PJT $(T, r, \gamma, \pi)$ is
  $\mathtt{width}(T) \coloneqq \max_{n \in \mathcal{V}(T)} \mathtt{size}(n)$.
\end{definition}

@@

\begin{definition}
  Let $f, g\colon 2^X \to \mathbb{R}$ be pseudo-Boolean functions. Operations
  such as addition and multiplication are defined pointwise as
  \[
    (f+g)(Y) \coloneqq f(Y)+g(Y),
  \]
  and
  \[
    (fg)(Y) \coloneqq f(Y)g(Y)
  \]
  for all $Y \subseteq X$.
\end{definition}
% TODO: do I need to formally consider extending a pseudo-Boolean function to a
% bigger domain?

@@

\begin{definition} \label{def:projection}
  Let $X$ be a set. For any $x \in X$, \emph{projection} $\exists_x$ is an
  endomorphism $\exists_x\colon \mathbb{R}^{2^X} \to \mathbb{R}^{2^X}$ defined
  as
  \[
    \exists_xf = f|_{x=1} + f|_{x=0}
  \]
  for any $f\colon 2^X \to \mathbb{R}$.
\end{definition}
% TODO: gamma will have to point to ADDs instead (and not be a bijection)

\begin{lemma}
  Let $f,g\colon 2^X \to \mathbb{R}$ be pseudo-Boolean functions represented by
  ADDs with $n$ and $m$ nodes, respectively, and $x \in X$. Then $f+g$ and $fg$
  can be computed in $\mathcal{O}(mn)$ time, and $\exists_xf$ can be computed in
  $\mathcal{O}(n^2)$ time.
\end{lemma}
\begin{proof}
  Addition and multiplication are implemented by \texttt{apply} algorithm which
  takes $\mathcal{O}(mn)$ time \cite{DBLP:journals/tc/Bryant86}. Projection
  consists of two restrictions and an addition by \cref{def:projection}. The
  computational complexity of restriction is dominated by the reduction
  operation that transforms a decision diagram into a minimal canonical form
  \cite{DBLP:journals/tc/Bryant86}. While the original reduction algorithm had a
  $\mathcal{O}(n \log n)$ complexity, caching can reduce it to $\mathcal{O}(n)$
  \cite{somenzi1998cudd}. Either way, the complexity of projection is still
  $\mathcal{O}(n^2)$.
\end{proof}

\begin{definition}
  Let $(T, r, \gamma, \pi)$ be a PJT, and $X$ be the set of variables. The
  functionality of DPMC execution (i.e., Algorithm~1 in the ADDMC paper
  \cite{DBLP:conf/aaai/DudekPV20}) can be represented by $\delta(r)$, where
  $\delta\colon \mathcal{V}(T) \to \mathbb{R}^{2^X}$ is a recursive function
  defined as
  \[
    \delta(t) =
    \begin{cases}
      \gamma(t) & \text{if } t \in \mathcal{L}(T) \\
      \exists_{\pi(t)} \prod_{u \in \mathcal{C}(t)} \delta(u) & \text{otherwise.}
    \end{cases}
  \]
  The range of $\delta(r)$ then contains a single real number, i.e., the answer.
\end{definition}

\begin{lemma}
  The number of variables that can appear in the ADDs within a PJT node is $\le
  k+1$.
\end{lemma}
\begin{proof}
  All such variables are introduced in a `clause' below and projected either in
  this node or in one of its ancestors. Each such variable would be in the bag
  of each clause that introduced it. By the tree decomposition properties, it
  will also be in the bag of both the projection node and the current node
  because they are on the path between the clause nodes (see
  \cref{fig:example_pjt}).
  \todo[inline]{How do I formalise this argument?}
\end{proof}

\begin{theorem}
  \textsc{DPMC} is FPT w.r.t. the PJT width. More specifically,
  \textsc{DPMC} running time is $\mathcal{O}(4^km(n+k))$, where $k$ is the
  width, $n$ is the number of clauses, and $m$ is the number of variables.
\end{theorem}
\begin{proof} % TODO: finish
  \begin{itemize}
  \item An ADD with $k$ variables can have up to $2^0 + 2^1 + \dots + 2^k =
    2^{k+1} - 1 = \mathcal{O}(2^k)$ nodes (including leaves).
  \item Multiplying $m$ ADDs can then take up to $(m-1)(2^{k+1}-1)^2 =
    \mathcal{O}(m4^k)$ (since one multiplication takes up to $(2^{k+1}-1)^2$
    time and the and the result will have up to $2^{k+1} - 1$ nodes because the
    domain stays the same).
  \item Each projection consists of two linear operations and a quadratic
    operation, so projecting $m$ variables will take $\mathcal{O}(4^km)$ time.
  \item In total, operations on a PJT node with $m$ ADDs, $k$ variables, $n$ of
    which are projected takes $\mathcal{O}(4^k(m+n))$ time.
  \item The total time taken by the DPMC execution stage is the sum of the time
    taken `inside' each PJT node. The number of such nodes is upper bounded by
    $m$, so the total time complexity of DPMC is $\mathcal{O}(4^km(n+k))$.
  \end{itemize}
\end{proof}

>>>>

\paragraph{Notation.} For any propositional formula $\phi$ and $p, q \in
\mathbb{R}$, let $[\phi]^p_q\colon 2^X \to \mathbb{R}$ be a
pseudo-Boolean function defined as
\[
  [\phi]^p_q(Y) \coloneqq
  \begin{cases}
    p & \text{if } Y \models \phi \\
    q & \text{otherwise}
  \end{cases}
\]
for any $Y \subseteq X$.

\begin{lemma} \label{lemma:add_size}
  An ADD with $n$ variables has $\mathcal{O}(2^n)$ nodes.
\end{lemma}
\begin{proof}
  An ADD with $n$ variables can be, at most, a complete binary tree of height
  $n+1$ (as measured by the number of vertices). It would then have $2^0 + 2^1 +
  \dots + 2^n = 2^{n=1} = \mathcal{O}(2^n)$ nodes.
\end{proof}

\begin{lemma}
  Let $\phi$ be a conjunction of $n$ literals. Then the ADD representation of
  $[\phi]^p_q$ can be constructed in $\mathcal{O}(2^n)$ time for any $p, q \in
  \mathbb{R}$ such that $p \ne q$.
\end{lemma}
\begin{proof}
  The ADD representation of $\phi$ itself can be constructed with a sequence of
  $n-1$ calls to \texttt{apply} with one of the two operands in each call always
  a literal. The number of variables in the other operand then follows the
  sequence $1, 2, 3, \dots, n-1$. By \cref{lemma:add_size}, the numbers of nodes
  in the ADD representations of these operands is then $\mathcal{O}(2^1),
  \mathcal{O}(2^2), \dots, \mathcal{O}(2^{n-1})$. Since one of the operands is
  of constant size, the overall time complexity of all calls to \texttt{apply}
  is then
  \[
    \mathcal{O}(2^1) + \mathcal{O}(2^2) + \dots + \mathcal{O}(2^{n-1}) =
    \mathcal{O}(2^n).
  \]

  Let $\alpha$ be the ADD representation of $\phi$. Then the ADD representation
  of $[\phi]_q^p$ is $(p-q)\alpha + q$. As scalar operations can obviously be
  implemented in linear time, the overall complexity remains $\mathcal{O}(2^n)$.
\end{proof}
% TODO: will need to define scalar operations on ADDs

\paragraph{References.}
\begin{itemize}
\item Related work without publicly available implementations:
  \begin{itemize}
  \item direct compilation to SDDs \cite{DBLP:conf/ecsqaru/ChoiKD13}
  \item direct compilation to PSDDs, also eliminating parameter variables (a thesis)
  \item maybe two more papers
  \end{itemize}
\end{itemize}

@@

\paragraph{TODO}
\begin{itemize}
\item do a literature search focused around these papers
\item use mathcal on V, L, E (and introduce them)
\item maybe use texttt for DPMC?
\item do I need to formally consider extending a pseudo-Boolean function to a
  bigger domain?
\item Define:
  \begin{itemize}
  \item an ADD as a DAG.
  \item What does it mean for an ADD to `have' variables? Maybe refer to
    pseudo-Boolean function sensitivity.
  \item Formal definition of a previous WMC instance (CNF, literal weight
    function) and the new definition (set of $2^{X} \to \mathbb{R}_{\ge 0}$
    pseudo-Boolean functions and a constant). Note that the constant idea is
    borrowed from \texttt{bklm16}.
  \item Boolean formula in CNF (perhaps this is too trivial to define),
  \item primal graph of a CNF formula (a.k.a.
    Gaifman/(variable) interaction/connectivity/clique/representing graph),
  \item Bayesian network (I already have a definition of these last two),
  \item moralisation of a Bayesian network (or of any DAG),
  \end{itemize}
\end{itemize}

@@

\begin{definition}[\cite{DBLP:conf/cp/DudekPV20}, verbatim but with one change]
  Let $X$ be a set of Boolean variables and $\phi$ be a CNF formula over $X$. A
  \emph{project-join tree} (PJT) of $\phi$ is a tuple $(T, r, \gamma, \pi)$
  where:
  \begin{itemize}
  \item $T$ is a tree with root $r \in \mathcal{V}(T)$,
  \item $\gamma\colon \mathcal{L}(T) \to \mathbb{R}^{2^X}$ maps leaves of $T$ to
    pseudo-Boolean functions, and
  \item $\pi\colon \mathcal{V}(T) \setminus \mathcal{L}(T) \to 2^X$ is a
    labelling function on internal nodes.
  \end{itemize}
  Moreover, $(T, r, \gamma, \pi)$ must satisfy the following two properties:
  \begin{enumerate}
  \item $\{\pi(n) : n \in \mathcal{V}(T) \setminus \mathcal{L}(T)\}$ is a
    partition of $X$, and
  \item for each internal node $n \in \mathcal{V}(T) \setminus \mathcal{L}(T)$,
    variable $x \in \pi(n)$, and clause $c \in \phi$ such that $x$ appears in
    $c$, the leaf node $\gamma^{-1}(c)$ must be a descendant of $n$ in $T$.
  \end{enumerate}
\end{definition}

Note that while Dudek et al. \cite{DBLP:conf/cp/DudekPV20} define $\gamma$ to be
a bijection between leaves of a PJT and clauses, our definition is more
general.

\begin{definition}[\cite{DBLP:conf/cp/DudekPV20}]
  \[
    \mathtt{Vars}(n) \coloneqq
    \begin{cases}
      \mathtt{Vars}(\gamma(n)) & \text{if } n \in \mathcal{L}(T) \\
      \left( \bigcup_{o \in \mathcal{C}(n)} \mathtt{Vars}(o) \right) \setminus
      \pi(n) & \text{if } n \not\in \mathcal{L}(T).
    \end{cases}
  \]
  \[
    \mathtt{size}(n) \coloneqq
    \begin{cases}
      |\mathtt{Vars}(n)| & \text{if } n \in \mathcal{L}(T) \\
      |\mathtt{Vars}(n) \cup \pi(n)| & \text{if } n \not\in \mathcal{L}(T).
    \end{cases}
  \]
  (Note that $\mathtt{size}(n)$ is the number of variables that can appear
  during the computation of $\delta(n)$ (excluding recursive calls).)
  The \emph{width} of a PJT $(T, r, \gamma, \pi)$ is
  $\mathtt{width}(T) \coloneqq \max_{n \in \mathcal{V}(T)} \mathtt{size}(n)$.
\end{definition}

@@

\begin{definition}
  Let $f, g\colon 2^X \to \mathbb{R}$ be pseudo-Boolean functions. Operations
  such as addition and multiplication are defined pointwise as
  \[
    (f+g)(Y) \coloneqq f(Y)+g(Y),
  \]
  and
  \[
    (fg)(Y) \coloneqq f(Y)g(Y)
  \]
  for all $Y \subseteq X$.
\end{definition}

@@

\begin{definition} \label{def:projection}
  Let $X$ be a set. For any $x \in X$, \emph{projection} $\exists_x$ is an
  endomorphism $\exists_x\colon \mathbb{R}^{2^X} \to \mathbb{R}^{2^X}$ defined
  as
  \[
    \exists_xf = f|_{x=1} + f|_{x=0}
  \]
  for any $f\colon 2^X \to \mathbb{R}$.
\end{definition}

\begin{lemma} \label{lemma:basic_ops}
  Let $f,g\colon 2^X \to \mathbb{R}$ be pseudo-Boolean functions represented by
  ADDs with $n$ and $m$ nodes, respectively, and $x \in X$. Then $f+g$ and $fg$
  can be computed in $\mathcal{O}(mn)$ time, and $\exists_xf$ can be computed in
  $\mathcal{O}(n^2)$ time.
\end{lemma}
\begin{proof}
  Addition and multiplication are implemented by \texttt{apply} algorithm which
  takes $\mathcal{O}(mn)$ time \cite{DBLP:journals/tc/Bryant86}. Projection
  consists of two restrictions and an addition by \cref{def:projection}. The
  computational complexity of restriction is dominated by the reduction
  operation that transforms a decision diagram into a minimal canonical form
  \cite{DBLP:journals/tc/Bryant86}. While the original reduction algorithm had a
  $\mathcal{O}(n \log n)$ complexity, caching can reduce it to $\mathcal{O}(n)$
  \cite{somenzi1998cudd}. Either way, the complexity of projection is still
  $\mathcal{O}(n^2)$.
\end{proof}

% NOTE: this is a rehash of equation (2) from the DPMC paper
\begin{definition} \label{def:abstraction}
  Let $(T, r, \gamma, \pi)$ be a PJT, and $X$ be the set of variables. The
  functionality of DPMC execution can be represented by $\delta(r)$, where
  $\delta\colon \mathcal{V}(T) \to \mathbb{R}^{2^X}$ is a recursive function
  defined as
  \begin{equation} \label{eq:execution}
    \delta(t) =
    \begin{cases}
      \gamma(t) & \text{if } t \in \mathcal{L}(T) \\
      \exists_{\pi(t)} \prod_{u \in \mathcal{C}(t)} \delta(u) & \text{otherwise.}
    \end{cases}
  \end{equation}
  The range of $\delta(r)$ then contains a single real number, i.e., the answer.
\end{definition}

% TODO: will need to rephrase this to fit my context
\begin{theorem}[Theorem~4 in \cite{DBLP:conf/cp/DudekPV20}, almost verbatim]
  Let $\phi$ be a CNF formula over a set $X$ of variables and $(S, \chi)$ be a
  tree decomposition of the primal graph of $\phi$ of width $w$. Then
  Algorithm~2 returns a PJT of $\phi$ of width at most $w+1$.
\end{theorem}

% \begin{lemma} % NOTE: this lemma is already proven above
%   Let $(S, \chi)$ be a tree decomposition, and $(T, r, \gamma, \pi)$ be its PJT
%   generated by Algorithm~2 in \cite{DBLP:conf/cp/DudekPV20}. For any $t \in T$,
%   the number of variables that appear in the ADDs within a PJT node is at
%   most $k+1$.
% \end{lemma}
% \begin{proof}
%   All such variables are introduced in a `clause' below and not projected in
%   descendants. Each such variable would be in the bag of each clause that
%   introduced it. By the tree decomposition properties, it will also be in the
%   bag of both the projection node and the current node because they are on the
%   path between the clause nodes (see \cref{fig:example_pjt}).
% \end{proof}

\begin{theorem}
  Let $(T, r, \gamma, \pi)$ be a PJT of width $k$. Then \textsc{DPMC} execution
  is fixed-parameter tractable with respect to the $k$. Specifically,
  \textsc{DPMC} execution time complexity is $\mathcal{O}(4^km(n+k))$, where $n
  = |\mathcal{L}(T)|$ is the number of clauses/leaves, and $m$ is the number of
  variables.
\end{theorem}
% TODO: start here. Use theorem 4 to show that the number of variables is
% bounded by primal treewidth
\begin{proof}
  \begin{itemize}
  \item Let us consider the overall complexity of all multiplications and
    projections throughout the recursive calls of $\delta$ in
    \cref{def:abstraction}. Clearly, the overall complexity is the sum of the
    complexity of operations performed within each call to $\delta$.
  \item Let $t \in \mathcal{V}(T)$ be an arbitrary vertex of $T$. If $t$ is a
    leaf,
  \item The total time taken by the DPMC execution stage is the sum of the time
    taken `inside' each PJT node. The number of such nodes is upper bounded by
    $m$, so the total time complexity of DPMC is $\mathcal{O}(4^km(n+k))$.
  \item Multiplying $m$ ADDs can then take up to $(m-1)(2^{k+1}-1)^2 =
    \mathcal{O}(m4^k)$ (since one multiplication takes up to $(2^{k+1}-1)^2$
    time and the and the result will have up to $2^{k+1} - 1$ nodes because the
    domain stays the same).
  \item Then, by \cref{lemma:basic_ops}, projecting $m$ variables will take
    $\mathcal{O}(4^km)$ time.
  \item In total, operations on a PJT node with $m$ children, $k$ variables, $n$ of
    which are projected takes $\mathcal{O}(4^k(m+n))$ time.
  \end{itemize}
\end{proof}
